---
title: "COGs"
knit: neurotransmissionevolution::custom_knit
---
```{r include=FALSE}
knitr::opts_chunk$set(
   warning     = FALSE
  ,message     = FALSE
  ,cache       = 1
  ,autodep     = T
) 
library(tidyverse)
library(magrittr)
library(data.table)
library(stringi)
library(neurotransmissionevolution)

load("../data/gene_ids.rda")
load("../data/adhesion_genes.rda")
load("../data/string_eukaryotes.rda")
```

```{r include=FALSE}
download_if_missing("https://stringdb-static.org/download/COG.mappings.v11.0.txt.gz")

cogs <- fread(
  "download/COG.mappings.v11.0.txt.gz",
  header           = F,
  stringsAsFactors = F,
  skip             = 1,
  sep              = "\t",
  col.names        = c("taxid.string_id","cog_id"),
  select           = c(1,4),
  quote            = ""
)
```
```{r results="asis", echo=FALSE}
cogs_spec <- tribble(
  ~col_name,           ~col_type,  ~used,  ~ex,                                ~desc,
  "taxid.string_id",   "character", "yes", "9606.ENSP00000269305",             "STRING protein ID",
  "start_position",    "numeric",   "no",  "1",                                "residue where orthogroup mapping starts",
  "end_position",      "numeric",   "no",  "393",                              "residue where orthogroup mapping ends",
  "cog_id",            "character", "yes", "NOG08732",                         "orthologous group ID",
  "protein_annotation","character", "no",  "Cellular tumor antigen p53; [...]","protein description"
) %T>% print_dataframe_specification(
  "data-raw/download/COG.mappings.v11.0.txt.gz",
  "https://stringdb-static.org/download/COG.mappings.v11.0.txt.gz",
  "Orthologous groups (COGs, NOGs, KOGs) and their proteins.",
  "cogs",
  "scale_down"
)
```

```{r results="hide"}
# Spliting first column into taxid and string_id
separated_ids <- cogs %$% stri_split_fixed(taxid.string_id, pattern = ".", n = 2, simplify = T)

cogs[["taxid"]]     <- separated_ids[, 1]
cogs[["string_id"]] <- separated_ids[, 2]

# Freeing up some memory
rm(separated_ids)
gc()

# keeping only eukaryotes
cogs %<>% select(-taxid.string_id) %>% filter(taxid %in% string_eukaryotes[["taxid"]])
gc()

# Subsetting cogs of interest
gene_cogs <- cogs %>%
    filter(string_id %in% gene_ids[["string_id"]]) %>%
    select(-taxid) %>%
    group_by(string_id) %>%
    summarise(n = n(), cog_id = paste(cog_id, collapse = " / "))

# Subsetting adhesion cogs of interest
gene_cogs_extra <- cogs %>%
    filter(string_id %in% adhesion_genes[["string_id"]]) %>%
    select(-taxid) %>%
    group_by(string_id) %>%
    summarise(n = n(), cog_id = paste(cog_id, collapse = " / "))
```

### Proteins with multiple COGs  
Some proteins are occasionally assigned to more than one orthologous group (OG) by the COG algorithm. Despite infrequent, a decision needs to be made to deal with these exceptions. Since we aim to identify the most ancient archetype that has been vertically inherited to humans, it is reasonable to resolve these proteins to their most ancient OG (in the case of a human protein assigned to more than one OG). However, this strategy can sometimes be inappropriate for multidomain proteins. For instance, the SHANK proteins are assigned to two OG, of which the most ancient one is simply associated with ankyrin repeats. A similar case involves the PLA2G4 family and C2 domains, where the six members are assigned to more than one OG. We decided to resolve SHANK and PLA2G4 proteins to their most recent OGs.
```{r eval=FALSE}
gene_cogs %>% filter(n > 1)
```
```{r results="asis", echo=FALSE}
gene_cogs %>% filter(n > 1) %>%
  knitr::kable(caption = "Proteins assigned to more than one OG.", booktabs = TRUE, linesep = "") %>%
  kableExtra::kable_styling(position = "left", latex_options = c("striped", "HOLD_position"))
```
```{r eval=FALSE}
gene_cogs_extra %>% filter(n > 1)
```
```{r results="asis", echo=FALSE}
gene_cogs_extra %>% filter(n > 1) %>%
  knitr::kable(caption = "One Adhesion protein has been assigned to more than one OG.", booktabs = TRUE, linesep = "") %>%
  kableExtra::kable_styling(position = "left", latex_options = c("striped", "HOLD_position"))
```

Manually resolving them to their most recent OG (which was determined in a previous run).
```{r}
# Resolving main proteins
gene_cogs_resolved <- tribble(
       ~string_id,   ~cog_id,
"ENSP00000371886", "KOG1325",# JMJD7-PLA2G4B
"ENSP00000356436", "KOG1325",# PLA2G4A
"ENSP00000396045", "KOG1325",# PLA2G4B
"ENSP00000290472", "KOG1325",# PLA2G4D
"ENSP00000382434", "KOG1325",# PLA2G4E
"ENSP00000380442", "KOG1325",# PLA2G4F
"ENSP00000293441", "KOG4375",# SHANK1
"ENSP00000469689", "KOG4375",# SHANK2
)

# Removing unresolved cases and adding manual assignments
gene_cogs %<>%
  filter(n == 1) %>%
  select(-n) %>%
  bind_rows(gene_cogs_resolved)

# Resolving extra adhesion proteins
gene_cogs_extra_resolved <- tribble(
       ~string_id,   ~cog_id,
"ENSP00000359085", "KOG3512",# NTNG1
)

# Removing unresolved cases and adding manual assignments
gene_cogs_extra %<>%
  filter(n == 1) %>%
  select(-n) %>%
  bind_rows(gene_cogs_extra_resolved)
```

Exporting for analysis:
```{r results="hide"}
# Exporting for package use
usethis::use_data(cogs, overwrite = TRUE)
usethis::use_data(gene_cogs, overwrite = TRUE)
usethis::use_data(gene_cogs_extra, overwrite = TRUE)
```
