---
# layout: 12pt,review
linenumbers: true
title: "Ionotropic receptors as the driving force behind human synapse establishment"
subtitle: "Supplementary Material"
author: 
  - Lucas H. Viscardi
  - Danilo O. Imparato
  - Maria Cátira Bortolini
  - Rodrigo J. S. Dalmolin
abstract: |
  Model uncertainty and limited data are fundamental challenges to
  robust management of human intervention in a natural system.
  These challenges are acutely highlighted by concerns that
  many ecological systems may contain tipping points, such as
  Allee population sizes.  Before a collapse, we do not know
  where the tipping points lie, if they exist at all.  Hence,
  we know neither a complete model of the system dynamics nor
  do we have access to data in some large region of state-space
  where such a tipping point might exist.
  
keywords:
  - Evolution
  - Nervous system
  
# header-includes:
   # - \usepackage[table]{xcolor}
   # - \usepackage{docmute}
   # - \usepackage{float}
   # - \usepackage{booktabs}
   
header-includes:
  - \usepackage{booktabs}
  - \usepackage{longtable}
  - \usepackage{array}
  - \usepackage{multirow}
  - \usepackage{wrapfig}
  - \usepackage{float}
  - \usepackage{colortbl}
  - \usepackage{pdflscape}
  - \usepackage{tabu}
  - \usepackage{threeparttable}
  - \usepackage{threeparttablex}
  - \usepackage[normalem]{ulem}
  - \usepackage{makecell}
  - \usepackage{xcolor}

## rmarkdown render options
output:
  pdf_document:
    toc: yes
    toc_depth: 6
    highlight: zenburn
    # pandoc_args: "--highlight=breezedark"
    # fig_caption: true
    # template: components/elsarticle.latex
    keep_tex: false
---

<!--- Project structure -->
# Project structure
This is the title page

# Preprocessing
Preprocessing

<!--- Eukaryota species tree -->
## Eukaryota species tree
Explanation

### NCBI Taxonomy tree
Explanation

### Duplicated Genera

### Hybrid tree
Explanation

<!--- Kegg annotation -->
## Gene selection and annotation

<!--- Kegg annotation -->
## Gene selection and annotation

<!--- Neuroexclusivity -->
## Neuroexclusivity
Explanation

### Expression

### Pathways

<!--- COG data -->
## COG data

<!--- Network -->
## Network

# Analysis
Analysis

```{r setup}
library(magrittr)
library(neurotransmissionevolution)
```

\input{download_files.tex}

Test tab2
```{r}
tibble::tribble(
  ~col_name,   ~col_type, ~used, ~ex,                  ~desc,
  "string_id", "numeric", "yes", "9606.ENSP000000666", "This represents the string protein identifier.",
  "length",    "numeric", "yes", "666",                "This represents the protein size"
) %>% print_dataframe_specification("a", "b", "caption")
```

