---
title: "Abundance"
knit: neurotransmissionevolution::custom_knit
---
```{r include=FALSE}
knitr::opts_chunk$set(
   warning     = FALSE
  ,message     = FALSE
  ,cache       = 1
  ,autodep     = T
  # ,cache.extra = file.mtime("download/link_pathway_entrez.tsv")
)
```

Loading initial resources:
```{r}
# Data manipulation
library(tidyverse)
library(magrittr)

# Utils
library(neurotransmissionevolution)

# Packaged data
data(
   cogs
  ,gene_ids
  ,gene_cogs
  ,string_eukaryotes
  ,package = "neurotransmissionevolution"
)

# Fresh analysis data
cog_roots                   <- read_tsv("geneplast_roots.tsv",             col_types = "ci")
clade_names                 <- read_tsv("geneplast_clade_names.tsv",       col_types = "ic")
clade_taxids                <- read_tsv("geneplast_clade_taxids.tsv",      col_types = "ici")

# Collapsing similar functions
gene_annotation <- read_tsv("../data/gene_annotation.tsv", col_types = "cc") %>%
  mutate(annotation = case_when(
     grepl("clearance",   annotation) ~ "depletion"
    ,grepl("degradation", annotation) ~ "depletion"
    ,grepl("transport",   annotation) ~ "synthesis"
    ,TRUE ~ annotation
  ))
```

We start by setting up reusable data frames with useful metrics.
```{r}
# If a gene has more than 1 COG, select the most recent one.
gene_cogs %<>%
  inner_join(cog_roots) %>%
  group_by(string_id) %>%
  filter(root == min(root))

# The function of a COG is the function of its proteins
cog_annotation <- gene_ids %>%
  inner_join(gene_cogs) %>%
  inner_join(gene_annotation) %>%
  distinct(cog_id, annotation)

# Number of proteins in a COG in every species
cog_abundance_by_taxid <- cogs %>%
  filter(cog_id %in% gene_cogs[["cog_id"]]) %>%
  count(taxid, cog_id,  name = "abundance") %>%
  left_join(cog_annotation)

# Mapping species to clade info
ordered_species <- string_eukaryotes %>%
  select(taxid, ncbi_name) %>%
  left_join(clade_taxids) %>%
  left_join(clade_names, by = c("lca" = "root")) %>%
  mutate(
     ncbi_name  = fct_reorder(ncbi_name, -taxid_order)
    ,clade_name = fct_reorder(clade_name, -taxid_order)
  )

# Plotting colors
annotation_colors <- c(
   "depletion"             = "#F40000"
  ,"excitability"          = "#FFAB00"
  ,"receptor-associated"   = "#D6EE00"
  ,"ionotropic receptor"   = "#43FF1C"
  ,"metabotropic receptor" = "#18FFFF"
  ,"signaling"             = "#0091EA"
  ,"g-protein"             = "#0033ff"
  ,"synthesis"             = "#AA00FF"
  ,"vesicle"               = "#FF00AA"
)
```

The average orthogroup abundances are finally computed for each species according to the function of orthogroups.
```{r}
avg_abundance_by_function <- cog_abundance_by_taxid %>%
  group_by(taxid, annotation) %>%
  summarise(avg_abundance = mean(abundance)) %>%
  # Adding species and clade info
  left_join(ordered_species)
```

```{r include=FALSE}
theme_main <- theme(
     panel.spacing      = unit(2.5, "pt")
    ,strip.background   = element_blank()
    ,panel.grid.major.x = element_blank()
    ,panel.grid.major.y = element_line(size = 0.25, linetype = "dotted", color = "#E0E0E0")
    ,strip.text.x       = element_text(size = 6, angle = 90, hjust = 0, vjust = 0.5, color = "#757575")
    ,strip.text.y       = element_text(size = 8, angle = 0, hjust = 0, vjust = 0.5, color = "#757575")
    ,axis.title         = element_text(size = 15, color = "#424242")
    ,axis.ticks.x       = element_blank()
    ,axis.text.x        = element_blank()
    ,axis.text.y        = element_text(size = 5.5)
    ,legend.position    = "none"
  )

theme_supplementary <- theme(
     panel.grid.major.x = element_line(color = "#E0E0E0", size = 0.25, linetype = "dotted")
    ,panel.grid.major.y = element_blank()
    ,strip.text.y       = element_text(size = 7, angle = 0, hjust = 0, vjust = 0.5, color = "#757575")
    ,strip.text.x       = element_text(size = 7, angle = 90, hjust = 0, vjust = 0.5, color = "#757575")
    ,axis.title         = element_text(size = 12, color = "#424242")
    ,axis.ticks         = element_line(colour = "grey20")
    ,axis.text.y        = element_text(size = 6, angle = 0, hjust = 1, vjust = 0.5, color = "#757575")
    ,axis.text.x        = element_text(size = 6)
  )

theme_average <- theme(
     panel.spacing      = unit(1, "pt")
    ,axis.title         = element_text(color = "#424242")
    ,axis.text          = element_text(color = "#757575")
    ,axis.text.x        = element_text(size = 7, angle = -45, vjust = 0, hjust = 0)
    ,axis.text.y        = element_text(size = 5)
    ,strip.background   = element_blank()
    ,strip.text         = element_text(color = "#757575")
    ,strip.text.y       = element_text(angle = 0, hjust = 0, vjust = 0.5)
  )

theme_big <- theme(
     panel.spacing      = unit(0.5, "pt")
    ,panel.grid.major.x = element_line(size = 0.1, linetype = "dashed")
    ,panel.grid.major.y = element_blank()
    ,strip.background   = element_blank()
    ,strip.text.x       = element_text(size = 8, angle = 90, hjust = 0.5, vjust = 0)
    ,strip.text.y       = element_text(size = 8, angle = 0, hjust = 0, vjust = 0.5)
    ,axis.text.x        = element_text(size = 6, angle = 90, vjust = 0, hjust = 0)
    ,axis.text.y        = element_text(size = 4.5)
    ,axis.ticks         = element_line(size = 0.1)
  )
```

Plotting:
```{r  uncapped_abundances, fig.show="hide", fig.height=35, fig.width=7, fig.cap="Abundance values by species. Species are ordered like in Supplementary Figure 1."}
# This vertical line indicates the first metazoan (Mnemiopsis leidyi / Ctenophora)
metazoa_line <- geom_vline(
   xintercept = "Mnemiopsis leidyi"
  ,color      = "#FF0000"
  ,linetype   = "11"
  ,alpha      = 1
  ,size       = 0.25
)

# Facet labeller
annotation_labels <- c(
   "ionotropic receptor"   = "Ionotropic \nreceptor"
  ,"metabotropic receptor" = "Metabotropic \nreceptor"
  ,"receptor-associated"   = "Receptor-\nassociated"
  ,"signaling"             = "Signaling"
  ,"vesicle"               = "Vesicle"
  ,"synthesis"             = "Synthesis"
  ,"depletion"             = "Depletion"
  ,"g-protein"             = "G protein"
  ,"excitability"          = "Excitability"
)

# Custom tick function
tick_function <- function(x) {
  seq(x[2], 0, length.out = 3) %>% head(-1) %>% tail(-1) %>% { ceiling(./5)*5 }
}

# Capping abundance values based on metazoan mean
capped_abundance_by_function <- avg_abundance_by_function %>%
  # mutate(capped_abundance = ifelse(abundance >= 100, 100, abundance)) %>%
  group_by(annotation) %>%
  mutate(
     max_abundance = avg_abundance[lca <= 29] %>% { mean(.) + 3*sd(.) }
    ,avg_abundance = ifelse(avg_abundance >= max_abundance, pmin(max_abundance, 100), pmin(avg_abundance, 100))
  )

# Plotting
abundance_plot <- ggplot(capped_abundance_by_function) +
  # Geoms  ----------------
  metazoa_line +
  geom_bar(
     aes(x = ncbi_name, y = avg_abundance, fill = annotation, color = after_scale(darken(fill, 0.1)))
    ,stat = "identity"
  ) +
  # Labels  ---------------
  xlab("Species") +
  ylab("Average protein abundance in orthologous groups") +
  # Scales ----------------
  scale_y_continuous(breaks = tick_function, minor_breaks = NULL) +
  scale_fill_manual(values = annotation_colors %>% darken(0.1)) +
  # Styling ---------------
  facet_grid(
     annotation ~ clade_name
    ,scales   = "free"
    ,space    = "free"
    ,labeller = labeller(annotation = annotation_labels)
  ) +
  theme_classic() + 
  theme_main
ggsave("plots/fig5_raw.pdf", abundance_plot, width = 18, height = 6)

# Uncapped abundances for supplementary text
abundance_plot %+% avg_abundance_by_function +
  facet_grid(
     clade_name ~ annotation
    ,scales   = "free"
    ,space    = "free"
    ,labeller = labeller(annotation = annotation_labels)
  ) +
  coord_flip() + 
  theme_supplementary
```
\clearpage

\thispagestyle{empty}

\pdfpageheight=36.6in

\begin{figure}[p]

\caption{Average protein abundances in orthologous groups by species. Orthologous groups are faceted according to their functions.}\label{fig:uncapped_abundances}

{\centering \includegraphics[height=35in, width=7in]{figs/analysis.abundance.uncapped_abundances-1} }

\end{figure}

\clearpage

\pdfpageheight=11in

Species-specific average abundances, now averaged by clades:
```{r clade_average, fig.align="center", fig.height=4, fig.width=8, fig.cap="Species-specific average abundances, now averaged by clades."}
ggplot(avg_abundance_by_function) +
  geom_bar(
    aes(x = clade_name, y = avg_abundance, fill = annotation, color = after_scale(darken(fill, 0.1)))
    ,stat = "summary"
    ,fun  = "mean"
  ) +
  scale_y_continuous(breaks = tick_function, minor_breaks = NULL) +
  scale_fill_manual(values = annotation_colors, guide = "none") +
  facet_grid(
     annotation ~ .
    ,scales   = "free"
    ,space    = "free_y"
    ,labeller = labeller(annotation = sub("\n", "", annotation_labels))
  ) +
  xlab("Clades") +
  ylab("Average abundance averaged by clade") +
  theme_classic() + 
  theme_average
```

Plain protein abundance in single orthogroups:
```{r all_abundances, fig.show="hide", fig.height=35, fig.width=20, fig.cap="Number of proteins in each neurotransmission COGs, for every species. COGs associated with multiple neurotransmission functions are colored grey. COG0733 is associated with depletion/vesicle. KOG2301 is associated with signaling/excitability. COG0814 is associated with vesicle/synthesis."}
# Collapsing COGs with multiple functions
cog_abundance_collapsed <- cog_abundance_by_taxid %>%
  group_by(taxid, cog_id) %>%
  summarise(
     annotation = paste(annotation, collapse = "/")
    ,abundance  = unique(abundance)
  ) %>%
  ungroup %>%
  left_join(ordered_species) %>%
  arrange(annotation) %>%
  mutate(cog_id = fct_inorder(cog_id))

# Adding colors for such COGs
annotation_colors %<>% c(
   "vesicle/synthesis"      = "#808080"
  ,"depletion/vesicle"      = "#808080"
  ,"signaling/excitability" = "#808080"
)

ggplot(cog_abundance_collapsed) +
  metazoa_line + 
  geom_bar(aes(x = ncbi_name, y = abundance, fill = annotation), stat = "identity") +
  scale_fill_manual(values = annotation_colors %>% darken(0.1), guide = "none") +
  scale_y_continuous(breaks = tick_function, minor_breaks = NULL) +
  facet_grid(clade_name ~ cog_id, scales = "free", space = "free_y") +
  xlab("Species") +
  ylab("Protein abundance in orthologous groups") +
  coord_flip() +
  theme_big
```
\clearpage

\thispagestyle{empty}

\newgeometry{top=0.4in, right=0.4in, bottom=0.4in, left=0.4in}

\pdfpagewidth=20.8in \pdfpageheight=36.8in

\begin{figure}[p]

\caption{Number of proteins in each neurotransmission COGs, for every species. COGs associated with multiple neurotransmission functions are colored grey. COG0733 is associated with depletion/vesicle. KOG2301 is associated with signaling/excitability. COG0814 is associated with vesicle/synthesis.}\label{fig:all_abundances}

{\centering  \includegraphics[height=35in, width=20in]{figs/analysis.abundance.all_abundances-1} }

\end{figure}

\clearpage

\restoregeometry

\pdfpagewidth=8.5in \pdfpageheight=11in
