---
title: "Cnidarian ionotropic receptors expression"
knit: neurotransmissionevolution::custom_knit
---
```{r include=FALSE}
knitr::opts_chunk$set(
   warning     = FALSE
  ,message     = FALSE
  # ,cache       = 1
  # ,autodep     = T
)
```

Loading resources:
```{r}
library(tidyverse)
library(magrittr)
library(readxl)
library(RCurl)
library(neurotransmissionevolution)

# Packaged data
data(
   cogs
  ,gene_ids
  ,gene_cogs
  ,gene_pathways
  ,string_eukaryotes
  ,package = "neurotransmissionevolution"
)

# Collapsing similar functions
gene_annotation <- read_tsv("../data/gene_annotation.tsv", col_types = "cc") %>%
  mutate(annotation = case_when(
     grepl("clearance",   annotation) ~ "depletion"
    ,grepl("degradation", annotation) ~ "depletion"
    ,grepl("transport",   annotation) ~ "synthesis"
    ,TRUE ~ annotation
  ))
```

Loading Supplementary Table S4, which contains genes expressed in neurons of the cnidarian *Nematostella vectensis*. The excel document contains multiple sheets, each reporting the genes expressed in a single cluster of neuronal cells. In this first step, we load all sheets into a data.frame.
```{r results="hide", message=FALSE, warning=FALSE}
xlsx_filename <- "nematostella_neuronal_genes.xlsx"

download.file(
   url      = "https://ars.els-cdn.com/content/image/1-s2.0-S0092867418305968-mmc4.xlsx"
  ,destfile = xlsx_filename
  ,mode     = "wb"
)

# Loading xlsx into data.frame
nematostella_neuronal_genes <- xlsx_filename %>% excel_sheets %>% map_df(~ read_excel(xlsx_filename, .))

# We are only interested in gene names and blast best hits
nematostella_neuronal_genes %<>% select(string_name = ...1, best_hit = Hsap.BBH) %>% distinct

file.remove(xlsx_filename)
```

Next, we look for *Nematostella vectensis* genes that are associated with ionotropic receptor COGs in STRING. Then, we search for these ionotropic receptor orthologs in the supplementary table.
```{r}
# Gathering gene info
gene_info <- gene_ids %>%
  na.omit %>%
  left_join(gene_cogs) %>%
  left_join(gene_annotation)

# Ionotropic receptor cogs
ionotropic_cogs <- gene_info %>% filter(annotation == "ionotropic receptor") %>% pull(cog_id) %>% unique

# Nematostella taxid
nematostella_taxid <- string_eukaryotes %>% filter(string_name == "Nematostella vectensis") %>% pull(taxid)

# Finding Nematostella ionotropic receptor orthologs
nematostella_receptors <- cogs %>%
  filter(taxid == nematostella_taxid & cog_id %in% ionotropic_cogs) %>%
  select(-taxid)

# Querying STRING for gene names 
nematostella_api_ids <- postForm(
   uri         ="http://string-db.org/api/tsv/get_string_ids"
  ,identifiers = nematostella_receptors %>% pull(string_id) %>% paste0(collapse="%0d")
  ,echo_query  = "1"
  ,species     = nematostella_taxid
) %>%
  read_tsv %>%
  select(string_id = queryItem, string_name = preferredName)

expressed_receptors <- nematostella_api_ids %>%
  left_join(nematostella_neuronal_genes) %>%
  left_join(nematostella_receptors) %>%
  arrange(cog_id)
```

Found receptors:
```{r echo=FALSE}
expressed_receptors %>%
  filter(!is.na(best_hit)) %>%
  select("COG" = cog_id, "String ID" = string_id, "Gene name" = string_name, "Human best hit" = best_hit) %>%
  knitr::kable(
     caption = "Human ionotropic receptor orthologs expressed in Nematostella neurons, according to work done by Sebé-Pedrós et al, 2018."
    ,booktabs = TRUE
    ,longtable = T
    ,linesep = ""
  ) %>%
  kableExtra::column_spec(4, width = "30em") %>%
  kableExtra::kable_styling(position = "left", latex_options = c("striped", "HOLD_position"))
```

Unfound receptors:
```{r echo=FALSE}
expressed_receptors %>%
  filter(is.na(best_hit)) %>%
  select("COG" = cog_id, "String ID" = string_id, "Gene name" = string_name) %>%
  knitr::kable(
     caption = "Human ionotropic receptor orthologs not found to be expressed in Nematostella neurons, according to work done by Sebé-Pedrós et al, 2018."
    ,booktabs = TRUE
    ,longtable = T
    ,linesep = ""
  ) %>%
  kableExtra::kable_styling(position = "left", latex_options = c("striped"))
```

