---
title: "Neuroexclusivity"
knit: neurotransmissionevolution::custom_knit
---
```{r include=FALSE}
knitr::opts_chunk$set(
   warning     = FALSE
  ,message     = FALSE
  ,cache       = 1
  ,autodep     = T
)
```

Loading resources:
```{r}
library(tidyverse)
library(magrittr)

data(
   gene_ids
  ,gene_pathways
  ,gene_expression
  ,package = "neurotransmissionevolution"
)
```

### Expression neuroexclusivity
We start by averaging all `gene_expression` collected from the Expression Atlas by tissue (`tpm_avg`). The averaged expression is filtered for values greather than 0.5 TPM. This ensures further computations only account for tissues in which genes are actually expressed. Then, we add the manual tissue classification indicating which tissues are nervous or not (described in Preprocessing). The neuroexclusivity index of a gene is the sum of its `tpm_avg` values in nervous tissues divided by the sum its values in all tissues.
```{r}
tissue_classification <- read_tsv(
  file      = "../data/neuroexclusivity_classification_tissue.tsv"
 ,col_types = "ci"
)

# Averaging TPM expression by tissue
avg_by_tissue <- gene_expression %>%
  group_by(ensembl_id, tissue) %>%
  summarise(tpm_avg = mean(tpm)) %>%
  filter(tpm_avg >= 0.5) %>%
  left_join(tissue_classification)

# Measuring expression neuroexclusivity
expression_neuroexclusivity <- avg_by_tissue %>%
  group_by(ensembl_id) %>%
  summarise(expression_neuroexclusivity = sum(tpm_avg[is_nervous == 1])/sum(tpm_avg)) %>%
  write_tsv("neuroexclusivity_expression.tsv")
```

### Pathway neuroexclusivity
To find the pathway neuroexclusivity of a gene, we simply divide the number of nervous pathways by the number of all pathways it participates in.
```{r}
pathway_classification <- read_tsv(
   file      = "../data/neuroexclusivity_classification_pathway.tsv"
  ,col_types = "cci"
)

link_pathway_entrez <- read_tsv(
   file      = "../data-raw/download/link_pathway_entrez.tsv"
  ,col_names = c("entrez_id", "pathway_id")
  ,col_types = "cc"
)

# Removing "hsa:" prefix
link_pathway_entrez[["entrez_id"]] %<>% str_split_n("\\:", 2)

# Pathway data related to our genes of interest
selected_genes_pathways <- link_pathway_entrez %>%
  filter(entrez_id %in% gene_ids[["entrez_id"]]) %>%
  left_join(pathway_classification) %>%
  drop_na # Dropping general pathways

# Measuring pathway neuroexclusivity
pathway_neuroexclusivity <- selected_genes_pathways %>%
  group_by(entrez_id) %>%
  summarise(pathway_neuroexclusivity = sum(is_nervous)/length(is_nervous)) %>%
  write_tsv("neuroexclusivity_pathway.tsv")
```

